---
title: "PLAN 672 - HW2"
author: "Chris Samoray"
date: "9/0/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F)

library(tidyverse)
library(here)
library(leaflet)
library(widgetframe)
```

```{r, echo=F, warning=F, include=F}
# load NO2 & O3 data from 2021
no2.df <- read_csv(here("HW_2", "data", "daily_42602_2021.csv"))

o3.df <- read_csv(here("HW_2", "data", "daily_44201_2021.csv"))


##############
# clean data
##############

# what does data look like?
glimpse(no2.df) 
glimpse(o3.df) 

# function to clean datasets
aqi.fcn <- function(x){
  
# remove spaces in column names
names(x) <- gsub(" ", "_", names(x)) 

# create new column of state/co fips code
x$stcofips <- paste(formatC(x$State_Code, width = 2, flag = "0"), formatC(x$County_Code, width = 3, flag = "0"), sep = "") # flag searches for leading zeros

print(x$stcofips %>% unique() %>% length()) # how many counties included
print(unique(nchar(x$stcofips))) # all have fips with 5 numbers

# select columns of interest
x <- x %>% select(stcofips, State_Name, County_Name, CBSA_Name, Site_Num, Longitude, Latitude, Date_Local, AQI, Parameter_Name)

x <- x %>%  mutate(month = substr(Date_Local, 6, 7))
 
return(x)
}

no2.df <- aqi.fcn(no2.df) # clean no2
o3.df <- aqi.fcn(o3.df) # clean o3

# include state abbreviations for plots
state.df <- data.frame(state.abb, state.name)
state.df <- state.df %>%  rename(State_Name = state.name)

# join with air pollutant datasets
no2.df <- left_join(no2.df, state.df, by = "State_Name")
o3.df <- left_join(o3.df, state.df, by = "State_Name")

# PR is left out, so do it separately
no2.df$state.abb <- ifelse(no2.df$State_Name == "Puerto Rico", "PR", no2.df$state.abb)
o3.df$state.abb <- ifelse(o3.df$State_Name == "Puerto Rico", "PR", o3.df$state.abb)

```


```{r,echo=F, warning=F, message=F}
##############
# create summary dfs
##############

# identify all monitoring stations in US
# monitor numbers can repeat, so need to remove duplicates by fips and monitor
no2.unique <- no2.df[!duplicated(no2.df[c(1,5)]), ] 
o3.unique <- o3.df[!duplicated(o3.df[c(1,5)]), ] 

# for every monitor in each county, identify the max daily aqi
no2.group <- no2.df %>% group_by(stcofips, Date_Local) %>%        arrange(desc(AQI), .by_group = T) %>%
                          summarize(maxAQI = max(AQI),
                          Site_Num = first(Site_Num),
                          State_Name = first(State_Name),
                          State_abb = first(state.abb),
                          County_Name = first(County_Name),
                          CBSA_Name = first(CBSA_Name),
                          Longitude = first(Longitude),
                          Latitude = first(Latitude))

# select aqi monitor with highest monthly average based on daily max aqi in each fips 
aqi.month.avg.high <- function(x){
x <- x %>% group_by(stcofips, month, Site_Num) %>% 
                          summarise(aqi_month_avg = mean(AQI),
                              State_Name = first(State_Name), # first in group, so okay bc will be the same for same monitor
                              State_abb = first(state.abb),
                              County_Name = first(County_Name),
                              CBSA_Name = first(CBSA_Name),
                              Longitude = first(Longitude),
                              Latitude = first(Latitude)) %>%
                          arrange(desc(aqi_month_avg), .by_group = T) %>% # select high
                          summarise(aqi_month_avg = first(aqi_month_avg),
                                    Site_Num = first(Site_Num), # why need this, shouldn't it be in the group?
                              State_Name = first(State_Name),
                              State_abb = first(State_abb),
                              County_Name = first(County_Name),
                              CBSA_Name = first(CBSA_Name),
                              Longitude = first(Longitude),
                              Latitude = first(Latitude),.groups = "keep")
return(x)
}

no2.high.co <- aqi.month.avg.high(no2.df)
o3.high.co <- aqi.month.avg.high(o3.df)

# identify the average monthly max aqi for each state
aqi.st.mo.high <- function(x) {x %>% group_by(State_Name, month) %>%
                arrange(desc(aqi_month_avg), .by_group = T) %>% 
                    summarise(max_month_avg = max(aqi_month_avg),
                              Site_Num = first(Site_Num),
                              State_abb = first(State_abb),
                              County_Name = first(County_Name),
                              CBSA_Name = first(CBSA_Name),
                              Longitude = first(Longitude),
                              Latitude = first(Latitude))
}

no2.high.st <- aqi.st.mo.high(no2.high.co)
o3.high.st <- aqi.st.mo.high(o3.high.co)

```
# Air Pollutant Monitors in the United States

The United States monitors air pollution throughout the country. Instruments measuring specific air pollutants, such as NO2 and O3, are stationed in many U.S. cities. However, not every monitor measures every air pollutant. Additionally, the same monitor identification number is sometimes used for different monitor at different locations. There are also more monitors measuring O3 than NO2: 1232 O3 monitors and 462 NO2 monitors. The locations of these monitors is identified in the map below. 

```{r monthly avg maps, echo=F, warning=F, message=F}

labels.no2 <- sprintf(
  "Location: %s <br/>  Site Number: %s",
  paste(no2.unique$County_Name, no2.unique$state.abb,  sep=", "),
  prettyNum(no2.unique$Site_Num)) %>% 
  lapply(htmltools::HTML)

labels.o3<- sprintf(
  "Location: %s <br/>  Site Number: %s",
  paste(o3.unique$County_Name, o3.unique$state.abb,  sep=", "),
  prettyNum(o3.unique$Site_Num)) %>% 
  lapply(htmltools::HTML)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircles(data = no2.unique, group = "NO2 Monitors", color = "#E69F00", 
             highlightOptions = highlightOptions(color = "green", 
             weight = 2, bringToFront = TRUE),
             label = labels.no2,
             labelOptions = labelOptions(
             style = list("font-weight" = "normal", padding = "3px 8px"),
             textsize = "8px",
             direction = "auto")) %>%
    addCircles(data = o3.unique, group = "O3 Monitors", color = "#56B4E9", 
             highlightOptions = highlightOptions(color = "green", 
             weight = 2, bringToFront = TRUE),
             label = labels.o3,
             labelOptions = labelOptions(
             style = list("font-weight" = "normal", padding = "3px 8px"),
             textsize = "8px",
             direction = "auto")) %>%
  addLayersControl(
    overlayGroups = c("NO2 Monitors", "O3 Monitors" ),
    options = layersControlOptions(collapsed = FALSE))
```
# Characteristics of Pollutants by State

The intensity of NO2 and O3 pollutants varies throughout the year. Taking the monthly average of the max AQI reading for each during that month reveals opposite seasonal distributions for NO2 and O3. In each state, NO2 is observed at higher measurements during the winter months compared with summer months. Conversely, O3 is generally highest during the summer months.

```{r, monthly avg plots, echo=F, warning=F, fig.height=20, fig.width=25}
# average monthly max AQI by state
# NO2
ggplot() + 
  geom_point(data = no2.high.st, aes(x=month, y=max_month_avg), alpha=.5, show.legend = FALSE, size =4) +
  labs(title = "Monthly Average NO2 Max AQI by State",x = "Month", y = "State Average NO2 max AQI")+
  facet_wrap(~State_abb)+
  theme_bw()+
theme(strip.text = element_text(size=16),
      axis.text = element_text(size =14),
      axis.title = element_text(size = 24),
      plot.title = element_text(size = 28))

# O3
ggplot() + 
  geom_point(data = o3.high.st, aes(x=month, y=max_month_avg), alpha=.5, show.legend = FALSE, size = 4) +
  labs(title = "Monthly Average O3 Max AQI by State",x = "Month", y = "State Average NO2 max AQI")+
  facet_wrap(~State_abb)+
  theme_bw()+
theme(
      strip.text = element_text(size=16),
      axis.text = element_text(size= 18),
      axis.title = element_text(size = 24),
      plot.title = element_text(size = 28))
```

# Trends by County

When including all counties--rather than just a single max measurement--the seasonal patterns persist of high NO2 in the winter and high 03 in the summer. But some states have more data than others. For instance, Alabama only has one data point per month, indicating that any air pollutant monitors are contained in a single county, whereas California has many data points for each month. Although the NO3 data looks rather evenly distributed, a few peculiarities in the O3 data would be worth exploring further in the future. Arizona, for example, has a few obvious points that stretch far beyond the max AQI of the next nearest county. States like Maryland and Virginia are also interesting, with many counties recording low O3 in February but concentrated O3 measurements during the rest of the year. 

```{r, echo=F, warning=F, fig.height=20, fig.width=25}
# average monthly AQI by county
# NO3
ggplot() + 
  geom_point(data = no2.high.co, aes(x=month, y=aqi_month_avg), alpha=.5, show.legend = FALSE, size = 4) +
  labs(title = "Monthly Average NO2 Max AQI for by County", x = "Month", y = "County Average NO2 max AQI")+
  facet_wrap(~State_abb)+
  theme_bw()+
theme(
      strip.text = element_text(size=16),
      axis.text = element_text(size= 18),
      axis.title = element_text(size = 24),
      plot.title = element_text(size = 28))

# o3
ggplot() + 
  geom_point(data = o3.high.co, aes(x=month, y=aqi_month_avg), alpha=.5, show.legend = FALSE, size = 4) +
  labs(title = "Monthly Average NO2 Max AQI for by County", x = "Month", y = "County Average NO2 max AQI")+
  facet_wrap(~State_abb)+
  theme_bw()+
theme(
      strip.text = element_text(size=16),
      axis.text = element_text(size =18),
      axis.title = element_text(size = 24),
      plot.title = element_text(size = 28))
```

# Distribution NO2 max AQI 

In general, NO2 AQI is concentrated below 25 AQI for the majority of states. However, many states have max AQI readings that extend beyond 50 AQI. Notably, Kansas has an AQI greater than 125, a measurement that should be evaluated further.

```{r, echo=F, warning=F, fig.height=20, fig.width=25}
# data density for max AQI of all county sensors  
ggplot(data = no2.group, aes(x=State_abb, y=maxAQI)) + 
  geom_violin(show.legend = FALSE) +
  labs(title = "Data Concentration of Monthly Average NO2 Max AQI for All County Sensors", x = "State", y = "Avg NO2 max AQI",
       colour=NULL, shape=NULL) + 
  theme_bw()+
theme(axis.text = element_text(size =18),
      axis.title = element_text(size = 24),
      plot.title = element_text(size = 28))
```

At the CBSA level, 03 pollution is worst in the western United States. States such as California and Colorado appear multiple times on a list of 25 cities with poor air quality. For NO2, states in the west again appear on the list. Unlike with O3, cities in the east of the country also top the NO2 list, including cities in Massachusetts, Virginia, New Hampshire and Georgia. Unlike 03, however, none of the NO2 measurements are over 100, the cutoff for acceptable air quality.

```{r, echo=F, warning=F, fig.height=20, fig.width=25}
# idenfity CBSAs with highest max avg AQI
# o3
top.o3 <- o3.high.co %>% group_by(CBSA_Name) %>% arrange(desc(aqi_month_avg), .by_group = T) %>% 
  filter(aqi_month_avg == first(aqi_month_avg)) %>% # to have only 1 record from each CBSA
  arrange(desc(aqi_month_avg))

top.o3 <-  top.o3[1:25, ] # select first 25, which are max

top.o3$CBSA_Name <- ifelse(top.o3$County_Name == "Mariposa", "Mariposa, CA", top.o3$CBSA_Name) # Mariposa does not have a CBSA defined, so name it here

  ggplot(top.o3)+
  geom_col(aes(x=reorder(paste(CBSA_Name), aqi_month_avg), y=aqi_month_avg), show.legend = F) +
    coord_flip()+ 
  labs(x="City", y="Highest Monthly Avg Max AQI ", title ="Top 25 cities with O3 pollution")+
  theme_minimal()+
theme(axis.text = element_text(size =18),
      axis.title = element_text(size = 24),
      plot.title = element_text(size = 28))
  
# NO2
top.no2 <- no2.high.co %>% group_by(CBSA_Name) %>% arrange(desc(aqi_month_avg), .by_group = T) %>% 
  filter(aqi_month_avg == first(aqi_month_avg)) %>% # to have only 1 record from each CBSA
  arrange(desc(aqi_month_avg))

top.no2 <-  top.no2[1:25, ] # select first 25, which are max

  ggplot(top.no2)+
  geom_col(aes(x=reorder(paste(CBSA_Name), aqi_month_avg), y=aqi_month_avg), show.legend = F) +
    coord_flip()+ 
  labs(x="City", y="Highest Monthly Avg Max AQI ", title ="Top 25 cities with NO2 pollution")+
  theme_minimal()+
theme(axis.text = element_text(size =18),
      axis.title = element_text(size = 24),
      plot.title = element_text(size = 28))

```

# Conclusions

Overall, the data suggest that air pollution is worse in the west than the rest of the country. In particular, O3 pollution is of significant concern in western cities, such as Los Angeles, Phoenix, and Denver. Moreover, these same cities are on the list of worst cities for NO2 pollution. 

In the west, wildfire is a significant concern. Given that O3 is worse in the summer than the winter, a future area of study would be to compare the EPA air pollution data with wildfire data for states that have many air pollution monitors, such as California. Fire data might be accessible from the California government website.

Looking into NO2 pollution other areas of the country might also be an interesting area to explore further. If NO2 is higher in the winter than the summer, perhaps there is a relationship with using increased energy for heating during the winter months. 

A distribution of max NO2 and O3 reading for cities with monitors provides an overview of geographic areas to consider for further study.
```{r, echo=F, message=F}
# max annual NO2 & O3 AQI recording by state
max.no2.high.co <- no2.high.co %>% group_by(CBSA_Name) %>%
  filter(aqi_month_avg == max(aqi_month_avg)) 

max.o3.high.co <- o3.high.co %>% group_by(CBSA_Name) %>%
  filter(aqi_month_avg == max(aqi_month_avg)) 

labels.no2<- sprintf(
  "Location: %s <br/>  Site Number: %s <br> Max Montly Avg AQI: %s <br> Month: %s",
  paste(max.no2.high.co$CBSA_Name, max.no2.high.co$state.abb,  sep=", "),
  prettyNum(max.no2.high.co$Site_Num),
  prettyNum(max.no2.high.co$aqi_month_avg),
  prettyNum(max.no2.high.co$month)) %>% 
  lapply(htmltools::HTML)

labels.o3<- sprintf(
  "Location: %s <br/>  Site Number: %s <br> Max Montly Avg AQI: %s <br> Month: %s",
  paste(max.o3.high.co$CBSA_Name, max.o3.high.co$state.abb,  sep=", "),
  prettyNum(max.o3.high.co$Site_Num),
  prettyNum(max.o3.high.co$aqi_month_avg),
  prettyNum(max.o3.high.co$month)) %>% 
  lapply(htmltools::HTML)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircles(data = max.no2.high.co, group = "Monthly Avg Max NO2", color = "#E69F00", 
             highlightOptions = highlightOptions(color = "green", 
             weight = 2, bringToFront = TRUE),
             label = labels.no2,
             labelOptions = labelOptions(
             style = list("font-weight" = "normal", padding = "3px 8px"),
             textsize = "8px",
             direction = "auto")) %>%
  addCircles(data = max.o3.high.co, group = "Monthly Avg Max O3", color = "#56B4E9", 
             highlightOptions = highlightOptions(color = "green", 
             weight = 2, bringToFront = TRUE),
             label = labels.o3,
             labelOptions = labelOptions(
             style = list("font-weight" = "normal", padding = "3px 8px"),
             textsize = "8px",
             direction = "auto")) %>%
  addLayersControl(
    overlayGroups = c("Monthly Avg Max NO2", "Monthly Avg Max O3"),
    options = layersControlOptions(collapsed = FALSE))
```

